Mallows, Black-box combinatorial optimization, limited budget, ???

### 2020-10-20 paper/main.tex
Sections 2, 3

\usepackage{amsmath,amssymb}
%\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xspace}
%% Useful packages
\usepackage{graphicx}
\usepackage{pdflscape,adjustbox}
\usepackage{multirow}
\usepackage{booktabs} % For formal tables
\usepackage{enumitem}
\usepackage{afterpage,tabularx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[square,numbers,sort]{natbib}
% Fixes for using natbib and llncs
\makeatletter
% required for natbib to have "References" printed and as section*, not chapter*
\renewcommand\bibsection%
{\section*{\refname\@mkboth{\MakeUppercase{\refname}}{\MakeUppercase{\refname}}}}
\renewcommand\@biblabel[1]{#1.}
\def\bibfont{\small}
\makeatother

@@

Permutations are defined as bijections of the set $[n]$ integer onto itself. The set of all permutations of $n$ items is denoted as $S_n$ and has cardinality $n!$. We denote permutations with Greek letters with exception of the inverse permutation denoted as $e=1, 2, 3, \ldots,n$. We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi$ and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma_{-1}=e$ always holds.

Distributions over permutations are functions that assign a probability value to each of the permutations in $S_n$, $p(\sigma)\in[0,1]$ \cite{critchlow91}. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma$ as follows:

\begin{equation}
p(\sigma)=\frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi}
\end{equation}

with two parameters, $\theta$ and $\sigma_0$, 
where permutation $\sigma_0$ a reference permutation that has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and $\theta$, the dispersion parameter controls this decay. The distance $d(\sigma,\sigma_0)$ is the Kendall's-$\tau$ distance. The normalization constant $\psi$ can be easily computed for the Kendall's-$\tau$ distance as well as for the Hamming, Cayley and Ulam distance~\cite{Irurozki2016b}.

One of the most common problems related ti probability distributions is that of learning the maximum likelihood parameters  given a sample of data $S$. For the MM, this problem tranlates to learning $\theta$ and $\sigma_0$ that best describe a sample of permutations.

The learning process is divide in two stages: first, we estimate the central permutation of the distribution, $\hat\sigma0$ and, second, compute the dispersion parameter, $\hat\theta$.

The exact maximum likelihood estimation is computationally hard~\cite{Dwork:2001:RAM:371920.372165}. However, the approximate learning requires polynomilal computational time and is guarantied to obtain high quality parameters~\cite{Caragiannis2013,Coppersmith:2010}.

This process of sample $S$ is as follows: first, compute $\hat\sigma0$ with the Borda count algorithm. Borda orders the items $[n]$ by their \textit{Borda score} increasingly, where the Borda score $B(i)$ is the average of each position $i$, $B(i) \propto \sum_{t\in S}  \sigma_t(i)$. Second, the computation of $\theta$ is casted as a numerical optimization problem~\cite{Irurozki2016b}.

\subsection{Unbalanced Borda}\label{sec:uborda}
Recently, it has been proposed an extension of Borda by the name of uBorda. uBorda considers a sample of permutations $S$ along with a weight $w(\sigma)$ for each permutation $\sigma$~\cite{}. Intuitively it is equivalent to replicating the permutations in the sample proportionally to their weight. It has been used to learn a MM in an evolving preference context. In this paper, we propose to use uBorda where each permutation in the sample $\sigma\in S$ is weighted by its fitness function $w(\sigma)=f(\sigma)$.

@@

\section{Methods}

@@

\section{UMM }
In this section we describe our main contribution: a novel algorithm for the optimization of a black box function over permutation spaces. This algorithm is called Unbalanced Mallows model (UMM) since it a probabilistic method based on a well-known probability distribution for permutations: the Mallows model.

The high-level behavior of UMM is as follows. It starts with an \textit{small} sample of $m'$ permutations that are randomly generated. UMM also keeps a probability distribution of a transformation of the sample. Then, for $m$ iterations, it generates a random, promising candidate solution and includes it in the sample. Each time a new permutation is included its fitness is computed and the internal probabilistic model is updated.

The probabilistic model is a MM that assigns higher probability values to the permutations in which the fitness value is known or expected to be smaller.\footnote{We assume a minimization problem.} The main contribution of our approach is that the distribution does not represent the current sample considered by the algorithm but instead, represents a variation of it. This variation is abstractly equivalent to replicating in the sample the permutations inversely proportionally to their fitness values, i.e., there will be many copies of the good solutions and few of the bad ones. One key property of UMM is the computational complexity of working with the original sample and the transformed sample is the same, i.e., polynomial. However, the question os how many times should we replicate each permutation in the sample? UMM includes a parameter $\rho$ that controls this point and will be detailed in Section~\ref{sec:learning}

At each iteration of the algorithm the following steps take place: (1) randomly generate a permutation and include it in the sample and (2) evaluate the new permutation and update the probabilistic model. These steps correspond to two statistical operations: (1) sample a probability distribution and (2) learn a probability distribution of the transformation of the sample. Both are describe in the subsequent sections.

@@

As shown in Section~\ref{sec:backgroud}, MM can be approximately learnt in two steps (1) estimating the location parameter $\sigma_0$, usually done with Borda algorithm, and (2) computing the scale parameter $\theta$. Since the scale parameter is set automatically, we focus on learning the first step.

%In order to minimize the number of permutations in the sample (and consequently the number of function evaluations) our proposed UMM learns the distribution of a transformation of the sample rather than the sample. This transformation consists on replicating the best fitted permutations. However, this is done abstractly: 
The location parameter can be shown to be the permutation in $S_n$ that minimizes the sum of the distances to the permutations in the sample~\cite{gMallows}, which is usually approximated with the Borda algorithm. Since we are interested on biasing the location towards the regions of the space where the fitness function is smaller, we would like to replicate the permutations a number of times that is inversely proportionally to their fitness values.
We can simulate the replication of the best fitted individuals by using uBorda, described in Section~\ref{sec:uborda} and setting $w(\sigma)=\rho^{f(\sigma)}$. This generalization requires setting a value for parameter $\rho$. We show how to choose this value in the experimental section.

\subsection{Sampling step}
The Mallows model is defined by two parameters. The location $\sigma_0$ is the permutation with the largest probability value, and in our problem, the permutation that is supposed to be the global optimum. The scale parameter $\theta$ controls both the variance and the expected distance of a random permutation. Given these two parameters, sampling a MM has been studied in \cite{Collas}\cite{Irurozki2016b}\cite{Regenwetter2004} where the computational complexities are $O(n\log n)$ and $O(n^2)$.

At each iteration of UMM $\sigma_0$ is given by the permutation obtained as a solution in the previous learning step. The scale $\theta$, in the other hand, is set automatically by the algorithm to control the diversification and intensification trade off: At the first iteration it is set so that the expected distance equals  ${n\choose 4}$ (half of the distance of the uniform distribution), and then it linearly decrease until the last iteration where the expected distance equals 1. Similar approaches have been taken in different contexts~\cite{arza2019approaching}.

@@

\subsection{Comparison to standard probabilistic optimization approaches.}
UMM is a population based algorithm. There exists a wide body of research on population based probabilistic algorithms, where most prominent examples are Estimation of Distribution Algorithms (EDAs). Our approach drastically differs from EDAs since EDAs assume that function evaluations are cheap and they can do many of them. On the contrary, UMM minimizes the number of function evaluations. Moreover, UMM does not rely on an evolving sequence of samples (called generations in evolutionary approaches). Instead, UMM considers a single sample of permutations during the whole execution that is increased iteratively, with one permutation each iteration.

>>>>

\usepackage{amsmath,amssymb}
%\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xspace}
%% Useful packages
\usepackage{graphicx}
\usepackage{pdflscape,adjustbox}
\usepackage{multirow}
\usepackage[]{algorithm2e}
\usepackage{booktabs} % For formal tables
\usepackage{enumitem}
\usepackage{afterpage,tabularx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[square,numbers,sort]{natbib}
% Fixes for using natbib and llncs
\makeatletter
% required for natbib to have "References" printed and as section*, not chapter*
\renewcommand\bibsection%
{\section*{\refname\@mkboth{\MakeUppercase{\refname}}{\MakeUppercase{\refname}}}}
\renewcommand\@biblabel[1]{#1.}
\def\bibfont{\small}
\makeatother

@@

Permutations are defined as bijections of the set $[n]$ integer onto itself. The set of all permutations of $n$ items is denoted as $S_n$ and has cardinality $n!$. We denote permutations with Greek letters with exception of the identity permutation denoted as $e=1, 2, 3, \ldots,n$. We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi$ and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma_{-1}=e$ always holds.

Distributions over permutations~\cite{critchlow91} are functions that assign a probability value to each of the permutations in $S_n$, $p(\sigma)\in[0,1]$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma$ as follows:

\begin{equation}
p(\sigma)=\frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi}
\end{equation}
with two parameters, $\theta$ and $\sigma_0$, Permutation $\sigma_0$ a reference permutation that has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and $\theta$, the dispersion parameter controls this decay. The distance $d(\sigma,\sigma_0)$ is the Kendall's-$\tau$ distance. The normalization constant $\psi$ can be easily computed for the Kendall's-$\tau$ distance as well as for the Hamming, Cayley and Ulam distance~\cite{Irurozki2016b}.

The learning process is divide in two stages: first, we estimate the central permutation of the distribution, $\hat\sigma0$ and, second, compute the dispersion parameter, $\hat\theta$~\cite{Irurozki2016b}.

The exact maximum likelihood estimator (MLE) for $\sigma_0$ is given by the well-known Kemeny ranking~\cite{Dwork:2001:RAM:371920.372165}. Unfortunately,  obteinig such ranking is NP-hard~\cite{Dwork:2001:RAM:371920.372165}. An alternative to the Kemeny ranking is the Borda ranking~\cite{Ali2011}, which has several advantages. First, Borda requires polynomial computational time. Second, Borda as an estimator for the MM is guarantied to obtain high quality parameters~\cite{Caragiannis2013}. Finally, in a general optimization perspective, Borda error is bounded~\cite{Coppersmith:2010}.

The Borda ranking of sample $S$ is computed as follows: first, for each item $i$ compute its Borda score $B(i) =  \sum_{t\in S}  \sigma_t(i)$. Then, assign rank 1 to the item with the smallest Borda score, rank 2 to the item with the second smallest score and so on, i.e., order the items by Borda score increasingly.

\subsection{Unbalanced Borda}\label{sec:uborda}
In these lines we introduce the core for our probabilisitc algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{}. In uBorda, the score of item $i$ is defined as follows:

\begin{equation}
\begin{split}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i),
\end{split}
\label{eq:uborda_score}
\end{equation}
and, the same as Borda, orders the items increasingly by their score.

Intuitively, the uBorda ranking of sample $S$ is equivalent to the Borda ranking of $S'$, being $S'$ an extension of $S$ where each each $\sigma\in S$ has been replicated proportionally to its weight $w(\sigma)$. In this paper, we would like to replicate each permutation $\sigma$ a number of times that is proportional to $f(\sigma)$. Therefore, we will use $w(\sigma)=\rho^{f(\sigma)}$ for a $\rho\in(0,1)$.

By setting $\rho=1$ the original Borda is recovered. However, for $\rho<1$ uBorda ranking will be closer to the rankings in $S$ with smaller value of $f(\sigma)$. Interestingly, Borda is computed in polynomial time. Clearly, the choice of the weight depends on the relevant property of a ranking in a particular domain. For example, it has been shown to have nice theoretical properties for the online estimation of the parameters of a sample with concept drift.

@@

\subsection{Previous method for black box optimization in combinatorial spaces: CEGO}

@@

\section{Unbalanced Mallows model: probabilistic BB optimization}
In this section we describe our main contribution: a novel algorithm for the optimization of a black box function over permutation spaces. This algorithm is called Unbalanced Mallows model (UMM) since it a probabilistic method based on a well-known probability distribution for permutations: the Mallows model.

\begin{algorithm}[t]
 \KwData{$n$, $m$,}
 \KwResult{how to write algorithm with \LaTeX2e }
 $S$\;
 $F$\;
 $\hat\sigma_0$ = uBorda($S, F,\rho$)\;
 \For{$m$ \text{iterations}}{
  linear decrease on $\theta$\;
  $\sigma$ = sample from a $MM(\hat\sigma_0, \theta)$\;
  Append $\sigma$ and $f(\sigma)$ to $S$ and $F$\;
   $\hat\sigma_0$ = uBorda($S, F,\rho$)\;
 }
 \Return $S, F$
 \caption{Unbalanced Mallows model for black box optimization}
 \label{alg:umm}
\end{algorithm}

The high-level behavior of UMM is shown in Algorithm~\ref{alg:umm}. It starts with an \textit{small} sample of $m'$ permutations, $S=\{\sigma_1, \ldots,  \sigma_{m'}\}$ that are randomly generated. UMM also keeps the function evaluation of each solution, $F=\{f(\sigma_1), \ldots, f(\sigma_{m'})\}$. An probabilistic model is used internally to guide the search. Then, for $m$ iterations\EKHINE{esto es exactamente así o es $m-m'$?}, UMM generates a random, promising candidate solution $\sigma$ and includes it in the sample $S$ while $f(\sigma)$ in included in $F$. At each iteration, the internal probabilistic model is updated.

The probabilistic model is a MM that assigns higher probability values to the permutations in which the fitness value is known or expected to be smaller.\footnote{We assume a minimization problem.} The main contribution of our approach is that the distribution does not represent the current sample $S$ considered by the algorithm but instead, represents a variation of it. This variation is based on the uBorda algorithm in Section~\ref{sec:uBorda} and is abstractly equivalent to replicating in the sample the permutations inversely proportionally to their fitness values, i.e., there will be many copies of the good solutions and few of the bad ones. One key property of UMM is the computational complexity of working with the original sample and the transformed sample is the same, i.e., polynomial. However, the question os how many times should we replicate each permutation in the sample? UMM includes a parameter $\rho$ that controls this point and will be detailed in Section~\ref{sec:learning}.

At each iteration of the algorithm the following steps take place: (1) randomly generate a permutation distributed according to the internal model and include it in the sample and (2) evaluate the new permutation and update the probabilistic model. These steps correspond to two statistical operations: (1) sample a probability distribution and (2) learn a probability distribution of the transformation of the sample. Both are describe in the subsequent sections.

@@

As shown in Section~\ref{sec:backgroud}, MM can be approximately learnt in two steps (1) estimating the location parameter $\sigma_0$ and (2) computing the scale parameter $\theta$. Since the scale parameter is set automatically, we focus on learning the first step.

%In order to minimize the number of permutations in the sample (and consequently the number of function evaluations) our proposed UMM learns the distribution of a transformation of the sample rather than the sample. This transformation consists on replicating the best fitted permutations. However, this is done abstractly: 
The location parameter of a MM is usually approximated with the Borda algorithm. Since we are interested on biasing the model towards the regions of the space where the fitness function is smaller, we would like to replicate the permutations a number of times that is inversely proportionally to their fitness values.
We can simulate the replication of the best fitted individuals by using uBorda, described in Section~\ref{sec:uborda} and setting $w(\sigma)=\rho^{f(\sigma)}$. This generalization requires setting a value for parameter $\rho$. We show how to choose this value in the experimental section.

\subsection{Sampling step}
Sampling can be done efficiently~\cite{Irurozki2016b} given the parameters that describe the model, i.e., $\sigma_0$ and $\theta$. Given these two parameters, sampling a MM has been studied in \cite{Collas}\cite{Irurozki2016b}\cite{Regenwetter2004} where the computational complexities are $O(n\log n)$ and $O(n^2)$.

At each iteration of UMM $\sigma_0$ is the permutation obtained as a solution in the previous learning step,~\ref{sec:learning}. The scale $\theta$, in the other hand, is set automatically by the algorithm to control the diversification and intensification trade off: At the first iteration it is set so that the expected distance equals  ${n\choose 4}$ (half of the distance of the uniform distribution), and then it linearly decrease until the last iteration where the expected distance equals 1. Similar approaches consisting of changing the variance to control the exploration-explotation tradeoff have been taken in different contexts~\cite{arza2019approaching}.

@@

\subsection{Comparison to standard probabilistic optimization approaches.}
UMM is a population based algorithm. There exists a wide body of research on population based probabilistic algorithms, where most prominent examples are Estimation of Distribution Algorithms (EDAs). Our approach drastically differs from EDAs since EDAs assume that function evaluations are cheap and they can do many of them. On the contrary, UMM minimizes the number of function evaluations. Moreover, UMM does not rely on an evolving sequence of samples (called generations in evolutionary approaches). Instead, UMM considers a single sample of permutations during the whole execution that is increased iteratively, with one permutation each iteration. 
\ekhine{bayesian optim}{ff}
\EKHINE{subir a arxiv uBorda}
